Donor Responses to Civil Society Repression

### 2025-03-16 index.qmd
Add keywords

---
title: "Donor Responses to Civil Society Repression"
subtitle: "Chapter for the *Handbook of Political Control*"
date: 2025-01-23
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu

@@

@@

This watering down of NGO programming illustrates interesting interplay between the different layers of Earl and Braithwaite's [-@EarlBraithwaite:2022] model of repression. Both de jure anti-NGO laws and the de facto civil society regulatory environment represent layer 3 institutions designed to limit participation in institutional politics. These limits cause civil society actors at layer 4 to shift to safer strategies and curtail their programming, which in turn can lead to worse consequences downstream. Formal and informal limits on civil society could be one mechanism behind the fact that foreign democracy aid increases the probability of civil war onset in states with weaker and less institutionalized civil society [@BraithwaiteLicht:2020]. Amid violent level 7 conflict, civil society restrictions may prevent NGOs from intervening effectively to address humanitarian challenges arising from violent conflict. For instance, Executive Order 13224—issued by George W. Bush in 2001—banned all financial transactions with any individual or organization on a list of Specially Designated Global Terrorists [@eo13224journal]. However, many humanitarian aid organizations (layer 4 institutions), work in environments where even the smallest circumstantial evidence could be grounds for designating these NGOs as having provided material support or resources for terrorism [@Bhungalia:2023]. Thus, many nonprofits may choose to avoid working in conflict zones. For example, after the US listed Al Shabaab in Somalia as a terrorist group, it resulted in a 88% reduction in civil society aid to Somalia over the next two years, with devastating consequences for Somalis [@Hayes:2017].

Other layers in this model of repression also point toward future research on donor responses to NGO restrictions. @BraithwaiteLicht:2020 find that states that receive foreign democracy assistance have a lower probability of internal conflict when civil society is more institutionalized. It is likely that broader social institutionalization (e.g., strong formal social institutions [level 5] and robust protest mobilization [level 6]) protects against more violent repression. We do not yet know how these other levels of institutionalization interact with level 3 civil society restriction, nor how official aid donors might condition their responses to repression. For instance, are donor states less likely to tame their programming when there are popular protests or powerful labor unions that can supplement resistance to government action?

@@

While the examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed. To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, and local populations to address larger structural issues that exacerbate government crackdowns on civil society. This chapter demonstrates that the use of layer 3 controls---such as law and administrative repression---does not just have an impact on layer 4 actors such as civil society institutions. They also have far-reaching repercussions on efforts to address and resolve violent conflict (layer 7) and support for vulnerable communities, including minorities, queer groups, women, refugees, and asylum seekers (layer 2), who often depend on specialized nonprofit groups for services and legal advice that may cease to exist in the face of legal restrictions and anti-NGO rhetoric. Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.

>>>>

---
title: "Donor Responses to Civil Society Repression"
subtitle: "Chapter for the *Handbook of Political Control*"
date: 2025-03-16
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu

@@

keywords: 
  - foreign aid
  - NGOs
  - NGO crackdown
  - civil society repression
  - international philanthropy
  - charitable giving

@@

This watering down of NGO programming illustrates interesting interplay between the different layers of Earl and Braithwaite's [-@EarlBraithwaite:2022] model of repression. Both de jure anti-NGO laws and the de facto civil society regulatory environment represent layer 3 institutions designed to limit participation in institutional politics. These limits cause civil society actors at layer 4 to shift to safer strategies and curtail their programming, which in turn can lead to worse consequences downstream. Formal and informal limits on civil society could be one mechanism behind the fact that foreign democracy aid increases the probability of civil war onset in states with weaker and less institutionalized civil society [@BraithwaiteLicht:2020]. Amid violent layer 7 conflict, civil society restrictions may prevent NGOs from intervening effectively to address humanitarian challenges arising from violent conflict. For instance, Executive Order 13224—issued by George W. Bush in 2001—banned all financial transactions with any individual or organization on a list of Specially Designated Global Terrorists [@eo13224journal]. However, many humanitarian aid organizations (layer 4 institutions), work in environments where even the smallest circumstantial evidence could be grounds for designating these NGOs as having provided material support or resources for terrorism [@Bhungalia:2023]. Thus, many nonprofits may choose to avoid working in conflict zones. For example, after the US listed Al Shabaab in Somalia as a terrorist group, it resulted in a 88% reduction in civil society aid to Somalia over the next two years, with devastating consequences for Somalis [@Hayes:2017].

Other layers in this model of repression also point toward future research on donor responses to NGO restrictions. @BraithwaiteLicht:2020 find that states that receive foreign democracy assistance have a lower probability of internal conflict when civil society is more institutionalized. It is likely that broader social institutionalization (e.g., strong formal social institutions [layer 5] and robust protest mobilization [layer 6]) protects against more violent repression. We do not yet know how these other layers of institutionalization interact with layer 3 civil society restriction, nor how official aid donors might condition their responses to repression. For instance, are donor states less likely to tame their programming when there are popular protests or powerful labor unions that can supplement resistance to government action?

@@

While the examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed. To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, and local populations to address larger structural issues that exacerbate government crackdowns on civil society. This chapter demonstrates that the use of layer 3 controls---such as law and administrative repression---does not just have an impact on layer 4 actors such as civil society institutions. They also have far-reaching repercussions on efforts to address and resolve violent conflict (layer 7) and support for vulnerable communities, including minorities, queer groups, women, refugees, and asylum seekers (layer 2), who often depend on specialized nonprofit groups for services and legal advice that may cease to exist in the face of legal restrictions and anti-NGO rhetoric. Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion. Recent developments within Western democracies---including the second Trump administration's attacks on both the USAID and the U.S. nonprofit sector and European donors' increasing withdrawal from external democracy support---raise new questions about how civic groups and private donors and foundations can respond to a changing international landscape.
